---
title: "finalproject"
output: html_document
date: "2024-03-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

Importing datasets
```{r}
#day is a more comprehensive data set and can be used more inclusively vs hourly.

day <- read_csv("day.csv", show_col_types = FALSE)

#rename some variables
day <- rename(day, date = dteday, year = yr, month = mnth, weather_situation = weathersit, count = cnt, serial = instant)

day
```
This code chunk answers Further Researched Question #1:

Did any major weather events cause changes in the average bike rentals, and was it something that may affect our data/predictions.

```{r}
#Changing levels for clarity
day$weather_situation <- factor(day$weather_situation, levels = c(1, 2, 3), labels = c("good", "okay", "bad"))

#Find average based on weather situation first
average_rentals_by_weather <- day %>%
  group_by(weather_situation) %>%
  summarize(average_rentals = mean(count, na.rm = TRUE))

print(average_rentals_by_weather)

ggplot(average_rentals_by_weather, aes(x = weather_situation, y = average_rentals)) +
  geom_bar(stat = "identity") +
  labs(x = "Weather Situation", y = "Average Bike Rentals", 
       title = "Average Bike Rentals by Weather Situation")

#Plotting bike rentals over time
ggplot(day, aes(x = date, y = count, color = weather_situation)) +
  geom_line() +
  labs(x = "Date", y = "Number of Bike Rentals",
       title = "Daily Bike Rentals Over Time by Weather Situation")
```

Now analyze the year 2012 alone to see if there were specific weather occurences
```{r}
#get the year 2012 values alone
day_2012 <- day %>% 
  filter(year(date) == 2012)

day$weather_situation <- factor(day$weather_situation, levels = c(1, 2, 3), labels = c("good", "okay", "bad"))

average_rentals_by_weather_2012 <- day_2012 %>%
  group_by(weather_situation) %>%
  summarize(average_rentals = mean(count, na.rm = TRUE))

print(average_rentals_by_weather_2012)

ggplot(average_rentals_by_weather_2012, aes(x = weather_situation, y = average_rentals)) +
  geom_bar(stat = "identity") +
  labs(x = "Weather Situation", y = "Average Bike Rentals", 
       title = "Average Bike Rentals by Weather Situation in 2012")

ggplot(day_2012, aes(x = date, y = count, color = weather_situation)) +
  geom_line() +
  labs(x = "Date", y = "Number of Bike Rentals",
       title = "Daily Bike Rentals Over Time by Weather Situation in 2012")
```
Upon analyzing the data in 2012, there are two low points that can be identified around June and October/November. When the bike shares were at all time lows at this point when the weather was bad, we searched for a reason for this. According to the top 5 weather incidents of 2012, there was hurricane Sandy in late October and also the June 29th Derecho where weather temperatures were at an all time high. The data points line up with the weather occurences and explain how weather can impact the number of bike rentals.
