---
title: "finalproject"
output: html_document
date: "2024-03-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

Importing datasets
```{r}
#day is a more comprehensive data set and can be used more inclusively vs hourly.

day <- read_csv("day.csv", show_col_types = FALSE)

#rename some variables
day <- rename(day, date = dteday, year = yr, month = mnth, weather_situation = weathersit, count = cnt, serial = instant)

day
```
This code chunk answers Further Researched Question #1:

Did any major weather events cause changes in the average bike rentals, and was it something that may affect our data/predictions.

```{r}
#Changing levels for clarity
day$weather_situation <- factor(day$weather_situation, levels = c(1, 2, 3), labels = c("good", "okay", "bad"))

#Find average based on weather situation first
average_rentals_by_weather <- day %>%
  group_by(weather_situation) %>%
  summarize(average_rentals = mean(count, na.rm = TRUE))

print(average_rentals_by_weather)

ggplot(average_rentals_by_weather, aes(x = weather_situation, y = average_rentals)) +
  geom_bar(stat = "identity") +
  labs(x = "Weather Situation", y = "Average Bike Rentals", 
       title = "Average Bike Rentals by Weather Situation")

#Plotting bike rentals over time
ggplot(day, aes(x = date, y = count, color = weather_situation)) +
  geom_line() +
  labs(x = "Date", y = "Number of Bike Rentals",
       title = "Daily Bike Rentals Over Time by Weather Situation")
```

Now analyze the year 2012 alone to see if there were specific weather occurences
```{r}
#get the year 2012 values alone
day_2012 <- day %>% 
  filter(year(date) == 2012)

day$weather_situation <- factor(day$weather_situation, levels = c(1, 2, 3), labels = c("good", "okay", "bad"))

average_rentals_by_weather_2012 <- day_2012 %>%
  group_by(weather_situation) %>%
  summarize(average_rentals = mean(count, na.rm = TRUE))

print(average_rentals_by_weather_2012)

ggplot(average_rentals_by_weather_2012, aes(x = weather_situation, y = average_rentals)) +
  geom_bar(stat = "identity") +
  labs(x = "Weather Situation", y = "Average Bike Rentals", 
       title = "Average Bike Rentals by Weather Situation in 2012")

ggplot(day_2012, aes(x = date, y = count, color = weather_situation)) +
  geom_line() +
  labs(x = "Date", y = "Number of Bike Rentals",
       title = "Daily Bike Rentals Over Time by Weather Situation in 2012")
```

Upon analyzing the data in 2012, there are two low points that can be identified around June and October/November. When the bike shares were at all time lows at this point when the weather was bad, we searched for a reason for this. According to the top 5 weather incidents of 2012, there was hurricane Sandy in late October and also the June 29th Derecho where weather temperatures were at an all time high. The data points line up with the weather occurences and explain how weather can impact the number of bike rentals.


This code chunk answers Further Researched Question #2:

How was attendance of the National Cherry Blossom Festival in D.C. and did the yearly event cause a spike in casual rentals.

Importing cherry blossoms dataset
```{r}
cherry_blossoms <- read_csv("cherry-blossoms_filtered_2011-2013.csv", show_col_types = FALSE)

#rename some variables
cherry_blossoms <- rename(cherry_blossoms, peak_bloom = "Yoshino Peak Bloom Date", festival_start_date = "Festival Start Date", festival_duration = "Festival Duration")

#selecting festival start date and duration 
cherry_blossoms <- select(cherry_blossoms, Year, festival_start_date, festival_duration)

#reformating festival start date and the end date
cherry_blossoms <- cherry_blossoms %>%
  mutate(
    Year = as.numeric(Year),
    festival_start_date = as.Date(festival_start_date - 1, origin=paste0(Year - 1, "-12-31")),
    festival_end_date = festival_start_date + as.numeric(festival_duration) - 1
  )

cherry_blossoms

```

```{r}
day <- day %>%
  mutate(Year = year(date))

#Perform an inner_join to merge day and cherry_blossoms data
merged_data <- inner_join(cherry_blossoms, day, by = "Year")

#Filter date
festival_data <- merged_data %>%
  filter(date >= festival_start_date & date <= festival_end_date)

#Select needed variables
festival_bike_rentals <- festival_data %>%
  select(festival_start_date, festival_end_date, date, count)

#Summarize bike rentals by festival
total_bike_rentals_by_festival <- festival_bike_rentals %>%
  group_by(festival_start_date, festival_end_date) %>%
  summarize(total_rentals = sum(count))

total_bike_rentals_by_festival

#calculates the mean bike rentals during festivals from 2011-2012
average_rentals_during_festivals <- merged_data %>%
  filter(date >= festival_start_date & date <= festival_end_date) %>%
  summarise(average_rentals_during_festival = mean(count, na.rm = TRUE))

average_rentals_during_festivals

#creates an identifier for the festival periods
cherry_blossoms <- cherry_blossoms %>%
  mutate(festival_id = row_number())

#merge with day dataset to exclude the periods of festival
day_with_festival <- day %>%
  left_join(cherry_blossoms, by = "Year") %>%
  mutate(is_festival = date >= festival_start_date & date <= festival_end_date,
         is_festival = ifelse(is.na(is_festival), FALSE, is_festival))

#calculates the mean bike rentals excluding festival days from 2011-2012
average_rentals_excluding_festival <- day_with_festival %>%
  filter(!is_festival) %>%
  summarise(average_rentals_excluding_festival = mean(count, na.rm = TRUE))

average_rentals_excluding_festival

```

Upon analyzing the average rentals during festival and excluding festival in the years of 2011 and 2012, we calculated that there is an 8.42% change. This means that during cherry blossom festival duration there is an increased amount of bike rentals compared to normal times of the year. Another finding is that in 2012 there is way more total bike rentals compared to 2011 during the days of cherry blossom festival. This can be explained by looking into the weather situation which was investigated earlier. In March of 2012 the temperature was 10 degrees Celsius higher compared to previous years, so normally it would be in the 50s but in March of 2012 it was in the 70s reaching the 80s. This could mean that warmer weather may have explained the drastic increase in total bike rentals in 2012 compared to 2011.   

