---
title: "finalproject"
output: html_document
date: "2024-03-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
```

Importing datasets
```{r}
#day is a more comprehensive data set and can be used more inclusively vs hourly.

day <- read_csv("day.csv", show_col_types = FALSE)

#rename some variables
day <- rename(day, date = dteday, year = yr, month = mnth, weather_situation = weathersit, count = cnt, serial = instant)

day
```
This code chunk answers Further Researched Question #1:

Did any major weather events cause changes in the average bike rentals, and was it something that may affect our data/predictions.

```{r}
#Changing levels for clarity
day$weather_situation <- factor(day$weather_situation, levels = c(1, 2, 3), labels = c("good", "okay", "bad"))

#Find average based on weather situation first
average_rentals_by_weather <- day %>%
  group_by(weather_situation) %>%
  summarize(average_rentals = mean(count, na.rm = TRUE))

print(average_rentals_by_weather)

ggplot(average_rentals_by_weather, aes(x = weather_situation, y = average_rentals)) +
  geom_bar(stat = "identity") +
  labs(x = "Weather Situation", y = "Average Bike Rentals", 
       title = "Average Bike Rentals by Weather Situation")

#Plotting bike rentals over time
ggplot(day, aes(x = date, y = count, color = weather_situation)) +
  geom_line() +
  labs(x = "Date", y = "Number of Bike Rentals",
       title = "Daily Bike Rentals Over Time by Weather Situation")
```
We look at the year 2011 

```{r}
#get the year 2011 values alone
day_2011 <- day %>% 
  filter(year(date) == 2011)

average_rentals_by_weather_2011 <- day_2011 %>%
  group_by(weather_situation) %>%
  summarize(average_rentals = mean(count, na.rm = TRUE))

print(average_rentals_by_weather_2011)

ggplot(average_rentals_by_weather_2011, aes(x = weather_situation, y = average_rentals)) +
  geom_bar(stat = "identity") +
  labs(x = "Weather Situation", y = "Average Bike Rentals", 
       title = "Average Bike Rentals by Weather Situation in 2011")

ggplot(day_2011, aes(x = date, y = count, color = weather_situation)) +
  geom_line() +
  labs(x = "Date", y = "Number of Bike Rentals",
       title = "Daily Bike Rentals Over Time by Weather Situation in 2011")
```
Match specific dates

```{r}
# Create a data frame with specific dates and their corresponding weather event labels
weather_events <- data.frame(
  date = mdy(c("January 26, 2011", "September 4, 2011", "September 5, 2011", 
               "September 6, 2011", "September 7, 2011", "September 8, 2011", 
               "August 28, 2011", "April 27, 2011", "April 28, 2011")),
  event = c("Snowstorm", "Tropical Storm Lee", "Tropical Storm Lee", 
            "Tropical Storm Lee", "Tropical Storm Lee", "Tropical Storm Lee", 
            "Hurricane Irene", "Tornado Outbreak", "Tornado Outbreak")
)

# Merge the weather events with day_2011 to get the counts for those dates
highlighted_data <- left_join(day_2011, weather_events, by = "date")

# Filter out the rows where label is NA to use in geom_point and geom_text
filtered_highlighted_data <- highlighted_data %>% 
  filter(!is.na(event))

# Create the plot
p <- ggplot(day_2011, aes(x = date, y = count)) +
  geom_line(aes(color = weather_situation)) +  # Line for daily data
  geom_point(data = filtered_highlighted_data, aes(x = date, y = count, color = event), size = 2) +  # Points for events
  geom_text(data = filtered_highlighted_data, aes(x = date, y = count, label = event), 
            hjust = "inward", vjust = "inward", size = 3, check_overlap = TRUE, nudge_y = 200) +  # Labels for events
  labs(title = "Daily Bike Rentals and Weather Events in 2011", x = "Date", y = "Number of Bike Rentals") +
  scale_color_manual(values = c("good" = "green", "okay" = "orange", "bad" = "red", 
                                "Snowstorm" = "blue", "Tropical Storm Lee" = "purple", 
                                "Hurricane Irene" = "black", "Tornado Outbreak" = "brown")) +
  theme_minimal() +
  theme(plot.title = element_text(size = 20),
        axis.title = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Print the plot
print(p)

# Calculate the monthly averages for 2011
monthly_avg_2011 <- day_2011 %>%
  group_by(month = floor_date(date, "month")) %>%
  summarize(average_count = mean(count))

# Match the weather events with the first day of the month to join with monthly averages
weather_events$month <- floor_date(weather_events$date, "month")

# Now, join the monthly average data with the weather events to get the average count for each event
weather_events_avg_2011 <- left_join(weather_events, monthly_avg_2011, by = "month")

# Filter out rows with NA in average_count after joining
weather_events_avg_2011 <- weather_events_avg_2011 %>% 
  filter(!is.na(average_count))

# Create the plot for 2011 with monthly averages and weather events
p1 <- ggplot(monthly_avg_2011, aes(x = month, y = average_count)) +
  geom_line(color = "blue", size = 1) +
  geom_point(data = weather_events_avg_2011, aes(x = date, y = average_count, color = event), size = 2) +
  scale_color_manual(values = c("Snowstorm" = "blue", "Tropical Storm Lee" = "purple",
                                "Hurricane Irene" = "red", "Tornado Outbreak" = "brown")) +
  labs(title = "Monthly Average Bike Rentals with Weather Events in 2011",
       x = "Month", y = "Average Bike Rentals") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Print the plot for 2011
print(p1)
```


Looking at the year 2011 we can match exact dates to weather occurrences. On January 26, there is an all time low for the 26-27th for count of bike rentals, this is related to the rush hour snowstorm that occurs. There are 506 rentals and 431 respectively. In the month of July there was a heat wave and had the hottest days which explains why it has the month with highest rentals. Then in August there were hurricanes which impacted the rentals even though the weather was okay. 


Now analyze the year 2012 alone to see if there were specific weather occurences
```{r}
#get the year 2012 values alone
day_2012 <- day %>%
  filter(year(date) == 2012)

average_rentals_by_weather_2012 <- day_2012 %>%
  group_by(weather_situation) %>%
  summarize(average_rentals = mean(count, na.rm = TRUE))

print(average_rentals_by_weather_2012)

ggplot(average_rentals_by_weather_2012, aes(x = weather_situation, y = average_rentals)) +
  geom_bar(stat = "identity") +
  labs(x = "Weather Situation", y = "Average Bike Rentals", 
       title = "Average Bike Rentals by Weather Situation in 2012")

ggplot(day_2012, aes(x = date, y = count, color = weather_situation)) +
  geom_line() +
  labs(x = "Date", y = "Number of Bike Rentals",
       title = "Daily Bike Rentals Over Time by Weather Situation in 2012")
```

merge for 2012

```{r}
# Specific dates and labels for weather events in 2012
weather_events_2012 <- data.frame(
  date = c(mdy("June 1, 2012"), mdy("June 29, 2012"), mdy("October 29, 2012"), mdy("October 30, 2012")),
  event = c("Tornado Outbreak", "Derecho", "Hurricane Sandy", "Hurricane Sandy")
)

# Merge the weather events with day_2012 to get the counts for those dates
highlighted_data_2012 <- left_join(day_2012, weather_events_2012, by = "date")

# Filter out the rows without events for use in geom_point
events_data_2012 <- highlighted_data_2012 %>% filter(!is.na(event))

# Create the plot for 2012
p_2012 <- ggplot(day_2012, aes(x = date, y = count, group = weather_situation, color = weather_situation)) +
  geom_line() +  # Line for daily data
  geom_point(data = events_data_2012, aes(x = date, y = count, color = event), size = 3) +  # Colored points for events
  scale_color_manual(values = c("good" = "green", "okay" = "orange", "bad" = "red", 
                                "Tornado Outbreak" = "blue", "Derecho" = "cyan", 
                                "Hurricane Sandy" = "chocolate")) +
  labs(title = "Daily Bike Rentals and Weather Events in 2012", x = "Date", y = "Number of Bike Rentals") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 20),
        axis.title = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Print the plot
print(p_2012)

# Calculate the monthly averages for 2012
monthly_avg_2012 <- day_2012 %>%
  group_by(month = floor_date(date, "month")) %>%
  summarize(average_count = mean(count))

# Match the weather events with the first day of the month to join with monthly averages
weather_events_2012$month <- floor_date(weather_events_2012$date, "month")

# Now, join the monthly average data with the weather events to get the average count for each event
weather_events_avg_2012 <- left_join(weather_events_2012, monthly_avg_2012, by = "month")

# Filter out rows with NA in average_count after joining
weather_events_avg_2012 <- weather_events_avg_2012 %>% 
  filter(!is.na(average_count))

# Create the plot for 2012 with monthly averages and weather events
p2 <- ggplot(monthly_avg_2011, aes(x = month, y = average_count)) +
  geom_line(color = "blue", size = 1) +
  geom_point(data = weather_events_avg_2011, aes(x = date, y = average_count, color = event), size = 2) +
  scale_color_manual(values = c("Tornado Outbreak" = "blue", "Derecho" = "cyan", 
                                "Hurricane Sandy" = "chocolate")) +
  labs(title = "Monthly Average Bike Rentals with Weather Events in 2011",
       x = "Month", y = "Average Bike Rentals") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Print the plot for 2012
print(p2)
```


Upon analyzing the data in 2012, there are two low points that can be identified around June and October/November. When the bike shares were at all time lows at this point when the weather was bad, we searched for a reason for this. According to the top 5 weather incidents of 2012, there was hurricane Sandy in late October and also the June 29th Derecho where weather temperatures were at an all time high. The data points line up with the weather occurences and explain how weather can impact the number of bike rentals.


This code chunk answers Further Researched Question #2:

How was attendance of the National Cherry Blossom Festival in D.C. and did the yearly event cause a spike in casual rentals.

Importing cherry blossoms dataset
```{r}
cherry_blossoms <- read_csv("cherry-blossoms_filtered_2011-2013.csv", show_col_types = FALSE)

#rename some variables
cherry_blossoms <- rename(cherry_blossoms, peak_bloom = "Yoshino Peak Bloom Date", festival_start_date = "Festival Start Date", festival_duration = "Festival Duration")

#selecting festival start date and duration 
cherry_blossoms <- select(cherry_blossoms, Year, festival_start_date, festival_duration)

#reformating festival start date and the end date
cherry_blossoms <- cherry_blossoms %>%
  mutate(
    Year = as.numeric(Year),
    festival_start_date = as.Date(festival_start_date - 1, origin=paste0(Year - 1, "-12-31")),
    festival_end_date = festival_start_date + as.numeric(festival_duration) - 1
  )

cherry_blossoms

```

```{r}
day <- day %>%
  mutate(Year = year(date))

#Perform an inner_join to merge day and cherry_blossoms data
merged_data <- inner_join(cherry_blossoms, day, by = "Year")

#Filter date
festival_data <- merged_data %>%
  filter(date >= festival_start_date & date <= festival_end_date)

#Select needed variables
festival_bike_rentals <- festival_data %>%
  select(festival_start_date, festival_end_date, date, count)

#Summarize bike rentals by festival
total_bike_rentals_by_festival <- festival_bike_rentals %>%
  group_by(festival_start_date, festival_end_date) %>%
  summarize(total_rentals = sum(count))

total_bike_rentals_by_festival

#calculates the mean bike rentals during festivals from 2011-2012
average_rentals_during_festivals <- merged_data %>%
  filter(date >= festival_start_date & date <= festival_end_date) %>%
  summarise(average_rentals_during_festival = mean(count, na.rm = TRUE))

average_rentals_during_festivals

#creates an identifier for the festival periods
cherry_blossoms <- cherry_blossoms %>%
  mutate(festival_id = row_number())

#merge with day dataset to exclude the periods of festival
day_with_festival <- day %>%
  left_join(cherry_blossoms, by = "Year") %>%
  mutate(is_festival = date >= festival_start_date & date <= festival_end_date,
         is_festival = ifelse(is.na(is_festival), FALSE, is_festival))

#calculates the mean bike rentals excluding festival days from 2011-2012
average_rentals_excluding_festival <- day_with_festival %>%
  filter(!is_festival) %>%
  summarise(average_rentals_excluding_festival = mean(count, na.rm = TRUE))

average_rentals_excluding_festival

```

Upon analyzing the average rentals during festival and excluding festival in the years of 2011 and 2012, we calculated that there is an 8.42% change. This means that during cherry blossom festival duration there is an increased amount of bike rentals compared to normal times of the year. Another finding is that in 2012 there is way more total bike rentals compared to 2011 during the days of cherry blossom festival. This can be explained by looking into the weather situation which was investigated earlier. In March of 2012 the temperature was 10 degrees Celsius higher compared to previous years, so normally it would be in the 50s but in March of 2012 it was in the 70s reaching the 80s. This could mean that warmer weather may have explained the drastic increase in total bike rentals in 2012 compared to 2011.   

